Data Specialization for a Staged Lambda Calculus

### 2014-11-09 paper/figures/splitting-overview.tex
minor tweaks to figures
working on the semantics

\begin{figure*}

\textbf{\lang\ Evaluation:}

\hspace{2em}\bbone-Evaluation: $e\mathbin{\redonesym}[\xi;v]$, where $\xi$ is a \emph{residual table} and $v$ is a \emph{partial value}.

\hspace{2em}\bbtwo-Evaluation: $e\mathbin{\redtwosym}q$, where $q$ is a \emph{residual} in the unstaged language \langTwo

@@

\hspace{4em}$l.r$ is the \emph{resumer} (representating all stage~\bbtwo\ subcomputatiosn in $e$)

@@

\caption{Summary of \lang\ evaluation and splitting.}
\label{fig:termSplitTwo}
\end{figure*}

>>>>

\begin{figure}
\textbf{Languages:}
\begin{itemize}
\item \lang: two-staged lambda calculus with product, sum, and
  recursive types
\item \langmono: an unstaged lambda calculus with products.
\end{itemize}

\textbf{\lang\ Evaluation Relations:}
\begin{itemize}
\item 
\bbone-Evaluation: $e\mathbin{\redonesym}[\xi;v]$, where $\xi$ is a \emph{residual table} and $v$ is a \emph{partial value}.

\item
\bbtwo-Evaluation: $e\mathbin{\redtwosym}q$, where $q$ is a \emph{residual} in the unstaged language \langTwo

\item 
Table reification: \ur{Fill in...}
\end{itemize}

@@

@@

\caption{Summary of \lang\ evaluation and splitting.}
\label{fig:terminology}
\label{fig:termSplitSummary}
\end{figure}

### 2014-11-09 paper/semantics.tex
working on the semantics

As is the case in essentially all staged language, the dynamic
semantics evaluates a term to a residual (a term in the target
language \langmono) in two stages. In the first stage, all stage-1
terms are evaluated; in the second stage all stage-2 terms are
evaluated.  To eliminate duplication, our semantics lifts out each
stage-2 term nested inside of a stage-1 expressions, into table which
makes it possible for the stage-2 term, actually the corresponding
residual obtained by reducing all stage-1 subterms, to be referred to
by use a dynamically generated variable (a label).

Since stage-1 evaluation does not evaluate stage-2 terms and since
stage-2 terms may be nested inside stage-1 terms, stage-1 evaluation
cannot reduce \lang\ terms to values but only to {\em partial values},
which are stage~\bbone\ terms that have been evaluated except for
their stage-2 subterms, which can only be~\bbtwo\ variables wrapped in
$\next$ blocks.  \figref{validPvalues} defines partial values.

The dynamic semantics consist of two evaluation relations $\redonesym$
and $\redtwosym$, called {\em 1-evaluation} and {\em 2-evaluation},
that respectively evaluates stage-1 and stage-2 expressions.

\input{figures/dynamics}

The relation 1-evaluation maps a stage-1 term to a partial value and a
residual table.

 In the example above, 
\begin{lstlisting} 
1`(next {`2`yhat`1`},7)`
\end{lstlisting}
is a partial value.

In the example
above,
\begin{lstlisting} 
2`let zhat = yhat*yhat*7 in zhat`
\end{lstlisting}
is a residual.

The dynamics for \lang\ implements the stage \bbone\ evaluation algorithm
described above using three judgments: $\redonesym$ (evaluation), $\redtwosym$
(specialization), and the auxiliary $\reifysym$ (reification). 
Along with these are three judgments identifying forms of values:
$\pvalsym$ (partial values), $\ressym$ (residuals), 
and the auxiliary $\fcon$ (residual table).
These judgments are related as follows:
\begin{itemize}
\item Evaluation sends a stage~\bbone\ term to a partial value.
In order to avoid duplicating subterms, evaluation collects explicit
substitutions into a single {\em residual table} that is also the output of
evaluation.
\item Specialization sends a stage~\bbtwo\ term to a residual.
\item Reification sends a single residual and a residual table to another residual.  
\item Evaluation depends on specialization for $\next$ terms, which have
stage~\bbtwo\ subterms.
\item Specialization depends on evaluation for $\prev$ and $\pause$, which have
stage~\bbone\ subterms. The residual table produced by evaluation is also
reified here.
\end{itemize}

Note that partial values and residuals are both terms for which stage \bbone\
computation has completed; however, partial values are stage~\bbone\ terms,
while residuals are stage~\bbtwo\ terms.

The $\redonesym$ judgment, defined in \ref{fig:diaSemantics}, takes an stage
\bbone\ term%
\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
the type system ensures that those must occur under a $\next$. Consequently
$\redonesym$ will never directly encounter a variable.}
%
to a {\em residual table} $\xi$ and a partial value $v$. The residual
table implements the explicit substitutions in our example---it maps fresh
stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
residuals. For non-staging features of \lang, $\redonesym$ is essentially
standard call-by-value evaluation, and gathers the subterms' residual tables
into a single one.

>>>>

As is the case in essentially all staged language, the dynamic
semantics evaluates a term to a residual (a term in the target
language \langmono) in two stages. In the first stage, all stage-1
terms are evaluated by a relation, called {\em 1-evaluation}; in the
second stage all stage-2 terms are evaluated by a relation
$\redtwosym$, called {\em 2-evaluation}.  To eliminate duplication,
our semantics lifts out each stage-2 term nested inside of a stage-1
expressions, into a table which makes it possible for the stage-2
term---more precisely the corresponding residual obtained by reducing
all stage-1 subterms---to be referred to by use a dynamically
generated variable (a label).

Since stage-1 evaluation does not evaluate stage-2 terms and since
stage-2 terms may be nested inside stage-1 terms, 1-evaluation cannot
reduce \lang\ terms to values.  Instead 1-evaluation yields a {\em
  partial value}, which which is a stage~\bbone\ term that have been
evaluated except for its stage-2 subterms, which can only
be~\bbtwo\ variables wrapped in $\next$ blocks.  \figref{validPvalues}
defines partial values.

In order to avoid duplication of stage-2 computations, 1-evaluation
labels stage-2 computations by generating variable symbols for them as
needed and stores their residuals in a {\em residual table} keyed by
the variable.

The relation 1-evaluation, defined by $\redonesym$ judgment as shown
in \ref{fig:diaSemantics}, takes an stage \bbone\ term%
\footnote{ The input to $\redonesym$ may be open on stage \bbtwo\ variables, but
the type system ensures that those must occur under a $\next$. Consequently
$\redonesym$ will never directly encounter a variable.}
%
to a residual table $\xi$ and a partial value $v$.  For non-staging
features of \lang, $\redonesym$ is essentially standard call-by-value
evaluation but in addition gathers the subterms' residual tables into
a single table. \ur{The syntax of residual tables should be defined in
  the terminology figure} Residual table entries are created by the
evaluation of \next: 1-evaluation reduces a \next expression by
reducing the body of the \next with 2-evaluation to obtain a
residual $q$ and returns a residual table that maps a freshly
generated variable to the residual.  \ur{Variable should be fresh in
  the rule.}

The relation 2-evaluation, defined by $\redtwosym$ judgment, as shows
in \figref{diaSemantics}, takes a stage \bbtwo\ term and reduces it to
a residual.  Apart from the evaluation of \prev and \pause, the
evaluation is straightforward. To reduce a \prev term, 2-evaluation
reduces the body by 1-evaluation to obtain a residual table and a
partial value.  It then reifies the residual table and the partial
value into an residual term, by creating for each variable-residual
pair in the table a corresponding variable binding in the residual,
finally ending with the partial value.
\ur{There seems to be errors in the semantics in the handling of
  recursive types.  I don't see any reason to do any reductions here.
If you want to, the description above needs to be updated.}

...UMUT: UP TO HERE...

 In addition, 1-evaluation {\em reifies} the contents of
the residual table into a residual when changing from stage-2 to stage

 The residual
table implements the explicit substitutions in our example---it maps fresh
stage \bbtwo\ variables (which may appear inside $\next$ blocks in $v$) to
residuals.

In order to avoid duplicating subterms, evaluation collects explicit
substitutions into a single {\em residual table} that is also the output of
evaluation.

 the auxiliary $\reifysym$ (reification). 
Along with these are three judgments identifying forms of values:
$\pvalsym$ (partial values), $\ressym$ (residuals), 
and the auxiliary $\fcon$ (residual table).
These judgments are related as follows:
\begin{itemize}
\item Evaluation sends a stage~\bbone\ term to a partial value.

\item Specialization sends a stage~\bbtwo\ term to a residual.

\item Reification sends a single residual and a residual table to another residual.

\item Evaluation depends on specialization for $\next$ terms, which have
stage~\bbtwo\ subterms.
\item Specialization depends on evaluation for $\prev$ and $\pause$, which have
stage~\bbone\ subterms. The residual table produced by evaluation is also
reified here.
\end{itemize}
